# An Introduction to Dygraph in R

Yanyun Chen

```{r, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```


### These are the codes demonstrated in the video tutorial. The video tutorial can be found at https://www.youtube.com/watch?v=ewQdzu47yZs


***Data Source: FRED, https://fred.stlouisfed.org/categories/9?t=cpi&ob=pv&od=desc#***

```{r library}
library(readxl)
library(dygraphs)
library(dplyr)
```
### 1. The Basic

```{r intro}
cpi_data <- read_excel("cpi.xlsx", col_types = c("date", 
    "numeric", "numeric"))
colnames(cpi_data) <- c("date","us","china")
us_cpi <- ts(cpi_data$us, frequency=12, start=c(1999,9))
dygraph(us_cpi)
```


**Note**: In the video, fill is demonstrated. But filling time series graph is generally not recommended.

```{r selector,title,label}
dygraph(us_cpi, main = "CPI, not seasonally adjusted: All Items for United States") %>% 
  dySeries("V1", label = "United States CPI",color='red') %>%
  dyAxis("y", label = "Index (2015=100)", valueRange = c(50, 120)) %>%
  dyOptions(axisLineWidth = 1.5, 
            drawGrid = FALSE) %>%
  dyRangeSelector(height = 20) %>%
  dyLegend(show = "follow")
```
### 2.Multiple Series
```{r multilple}
cpi <- ts(cpi_data[,-1], frequency=12, start=c(1999,9))
dygraph(cpi)
```
```{r multilple/options}
dygraph(cpi,main="CPI, not Seasonally Adjusted: All Items") %>%
  dySeries("us", label = "United States",color='red') %>%
  dySeries("china", label = "China",color='blue') %>%
  dyAxis("y", label = "Index (2015=100)") %>%
  dyOptions(axisLineWidth = 1.5, drawGrid = TRUE) %>%
  dyRangeSelector(height = 20)
```

### 3.Events

**Note**: In the video, drawing event line along with shaded region is demonstrated. Another option is to draw annotation along with shaded region, which is demonstrated below.

```{r events}
dygraph(cpi, main="CPI, not Seasonally Adjusted: All Items") %>%
  dySeries("us", label = "United States") %>%
  dySeries("china", label = "China") %>%
  dyRangeSelector(height = 20) %>%
  dyEvent("2019-12-01", "COVID-19", labelLoc = "bottom") %>%
  dyAnnotation("2008-05-01", "2008 Financial Crisis", attachAtBottom = TRUE, width = 150) %>%
  #dyEvent("2008-05-01", "2008 Financial Crisis", labelLoc = "bottom") %>%
  dyShading(from = "2008-01-01", to = "2008-12-01")
```


### 4.Prediction
```{r upper/lower}
china_cpi <- ts(cpi_data$china, frequency=12, start=c(1999,9))
hw <- HoltWinters(china_cpi)
predicted <- predict(hw, n.ahead = 36, prediction.interval = TRUE)
all <- cbind(china_cpi, predicted)

dygraph(all, "CPI, not Seasonally Adjusted: All Items for China") %>%
  dySeries("china_cpi", label = "Actual") %>%
  dySeries(c("predicted.lwr", "predicted.fit", "predicted.upr"), label = "Predicted")
```


```{r upper/lower/more}
china_cpi_200 <- ts(cpi_data$china[1:200], frequency=12, start=c(1999,9))
hw <- HoltWinters(china_cpi_200)
predicted <- predict(hw, n.ahead = 64, prediction.interval = TRUE)
all <- cbind(china_cpi, predicted)

dygraph(all, "CPI, not Seasonally Adjusted: All Items for China") %>%
  dySeries("china_cpi", label = "Actual") %>%
  dySeries(c("predicted.lwr", "predicted.fit", "predicted.upr"), label = "Predicted")
```

