# Network Visualization in R

Yunze Pan

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(visNetwork)
library(igraph)
library(ggplot2)
library(tidyr)
library(plotly)
library(RColorBrewer)
library(reshape2)
```

## Introduction

`visNetwork` is a powerful tool in R to help us describe networks and explore the structure visually. It is extremely useful for us to obtain valuable information from an interactive network graph. In this tutorial, we will offer a quick introduction for newcomers to learn concepts of creating networks in R. Hope you will enjoy!

## Installation

The main packages we are going to use for network visualization in R are `visNetwork` and `igraph`. They can be installed with install.packages("visNetwork") and install.packages("igraph").

```{r, eval=FALSE}
library(visNetwork)
library(igraph)
```

## Dataframe

In this section we will create a small network that simulates student interactions on campus. Our objective is to get you familiar with using `visNetwork` as quickly as as possible. In order to visualize interactive networks, we will first read two datasets (a nodes data.frame and an edges data.frame). Then, we can explore the various layout options by adding different variables on our nodes data.frame and edges data.frame.

### Nodes

A nodes data.frame must include a `id` column. Each `id` represents the node we want to display in our graph. Other optional columns can also be added into our nodes data.frame. They can help us to distinguish nodes in our graph. For example, each node is a student with a unique assigned `id`, his/her `name`, `major`, and `major.type`.

```{r}
nodes <- data.frame(id=1:7, # id column (must be called id)
                    name=c("Asher","Bella","Chloe","Daniel","Emma","Frank","Gabriel"), # student names
                    major=c("CS","CS","CS","STAT","DS","DS","DS"), # CS: computer science major, STAT: statistics major, DS: data science major
                    major.type=c(1,1,1,2,3,3,3)) # 1: CS, 2: STAT, 3: DS
data.frame(nodes)
```

### Edges

An edges data.frame must include a `from` column and a `to` column denoting the starting node and ending node of each edge. We use `id` to represent the starting node and ending node. We also add a `weight` column on our edges data.frame to describe the frequency of interactions between two nodes. For example, in the first row, we know `student 1` reached out to `student 2` once.

```{r}
edges <- data.frame(from=c(1,1,2,3,5,5,6,7),
                    to=c(2,4,3,1,4,6,7,5),
                    weight=c(1,1,1,1,1,1,1,1))
data.frame(edges)
```

## Visualiztion

Now we can visualize our student interaction network using `visNetwork`. Examples are showed as below. We will start from the default setting and then move on to customize our network for a better interactive visualization.

### Minimal Example

```{r, fig.height=3, fig.width=9}
visNetwork(nodes, edges)
```

### Customize Node

```{r, fig.height=3, fig.width=9}
colors <- colorRampPalette(brewer.pal(3, "RdBu"))(3) # use three colors to distinguish students by their majors
nodes <- nodes %>% mutate(shape="dot", # "shape" variable: customize shape of nodes ("dot", "square", "triangle")
                          shadow=TRUE, # "shadow" variable: include/exclude shadow of nodes
                          title=major, # "title" variable: tooltip (html or character), when the mouse is above
                          label=name, # "label" variable: add labels on nodes
                          size=20, # "size" variable: set size of nodes
                          borderWidth=1, # "borderWidth" variable: set border width of nodes
                          color.background=colors[major.type], # "color.background" variable: set color of nodes
                          color.border="grey", # "color.border" variable: set frame color
                          color.highlight.background="yellow", # "color.highlight.background" variable: set color of the selected node
                          color.highlight.border="black") # "color.highlight.border" variable: set frame color of the selected node
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>% # "main" variable: add a title
  visLayout(randomSeed=4) # give a random seed manually so that the layout will be the same every time
```

### Customize Edge

```{r, fig.height=3, fig.width=9}
edges <- edges %>% mutate(width=weight*3, # "width" variable: set width of each edge
                          color="lightgrey", # "color" variable: set color of edges
                          arrows="to", # "arrows" variable: set arrow for each edge ("to", "middle", "from ")
                          smooth=TRUE) # "smooth" variable: each edge to be curved or not
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>% 
  visLayout(randomSeed=4)
```

### Add Legend Based on Groups

```{r, fig.height=3, fig.width=9}
nodes <- nodes %>% mutate(group=major) # add a "group" column on node data.frame and add groups on nodes
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>%
  visLayout(randomSeed=4) %>% 
  visGroups(groupname="CS", color=colors[1]) %>% # color "colors[1]" for "CS" group 
  visGroups(groupname="STAT", color=colors[2]) %>%
  visGroups(groupname="DS", color=colors[3]) %>%
  visLegend(width=0.1, position="right", main="Academic Major") # "position" variable: set position ("left", "right") 
```

### Select by Node

```{r, fig.height=3, fig.width=9}
nodes <- nodes %>% select(-group) # remove "group" column because we don't want to show legend this time
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>%
  visLayout(randomSeed=4) %>% 
  visOptions(nodesIdSelection=TRUE, # "nodesIdSelection" variable: select a node by id
             selectedBy="major") %>% # "selectedBy" variable: select a node by the values of a column such as "major" column
  visLegend()
```

### Highlight Nearest Nodes

```{r, fig.height=3, fig.width=9}
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>% 
  visLayout(randomSeed=4) %>% 
  visOptions(highlightNearest = list(enabled = TRUE, # "enabled" variable: highlight nearest nodes and edges by clicking on a node
                                     degree = 2)) # "degree" variable: set degree of depth
```

### Edit Network

```{r, fig.height=3, fig.width=9}
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>%
  visLayout(randomSeed=4) %>% 
  visOptions(highlightNearest=TRUE, # degree of depth = 1
             nodesIdSelection=TRUE,
             selectedBy="major",
             manipulation=TRUE) %>%  # "manipulation" variable: add/delete nodes/edges or change edges
  visLegend()
```

### Add Navigation Buttons and Control Interactions

```{r, fig.height=3, fig.width=9}
visNetwork(nodes, edges, width="100%", main="Student Interaction Network") %>%
  visLayout(randomSeed=4) %>%
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE,
             selectedBy="major") %>% 
  visInteraction(hideEdgesOnDrag=TRUE, # "hideEdgesOnDrag" variable: hide edges when dragging the view
                 dragNodes=TRUE, # "dragNodes" variable: hide nodes when dragging the view
                 dragView=TRUE, # "dragView" variable: enable or not the movement of the full network
                 zoomView=TRUE, # "zoomView" variable: enable or not the zoom (use mouse scroll)
                 navigationButtons=TRUE) %>% # "navigationButtons" variable: show navigation buttons
  visLegend()
```

## Export

Finally, we use `visSave()` to save network in html file.

```{r}
our_network <- visNetwork(nodes, edges)
visSave(our_network, file = "Student Interaction Network.html", background="white")
```

## Help?

More information about `visNetwork`.

```{r}
?visNodes
?visEdges
?visOptions
?visGroups
?visLegend
?visLayout
```

## Social Network Analysis

We have already learned how to visualize the interactive network. To help you better understand its application, we will use `visNetwork` and `igraph` to perform our social network analysis.

### Dataset

We will investigate interactions in the movie **Star Wars Episode IV**. First, we import two csv files ("nodes.csv" and "edges.csv"). Each node in "nodes.csv" is a character and each edge in "edges.csv" tells whether two characters appeared together in a scene of the movie. Thus, edges are undirected. Since characters may appear in multiple scenes together, each edge has a `weight`.

```{r}
sw_nodes <- read.csv("https://raw.githubusercontent.com/pablobarbera/data-science-workshop/master/sna/data/star-wars-network-nodes.csv")
head(sw_nodes)
```

```{r}
sw_edges <- read.csv("https://raw.githubusercontent.com/pablobarbera/data-science-workshop/master/sna/data/star-wars-network-edges.csv")
head(sw_edges)
```

We group our characters ("dark side" or "light side" or "other").

```{r}
dark_side <- c("DARTH VADER", "MOTTI", "TARKIN")
light_side <- c("R2-D2", "CHEWBACCA", "C-3PO", "LUKE", "CAMIE", "BIGGS", "LEIA", "BERU", "OWEN", "OBI-WAN", "HAN", "DODONNA", "GOLD LEADER", "WEDGE", "RED LEADER", "RED TEN", "GOLD FIVE")
other <- c("GREEDO", "JABBA")
sw_nodes$group <- NA
sw_nodes$group[sw_nodes$name %in% dark_side] <- "dark side"
sw_nodes$group[sw_nodes$name %in% light_side] <- "light side"
sw_nodes$group[sw_nodes$name %in% other] <- "other"
```

Let's try another network package called `igraph` to explore the network.

First, we use the `graph_from_data_frame` function, which needs two arguments: `d` and `vertices`. The `igraph` object `g` indicates that there are 22 nodes and 66 edges.

```{r}
g <- graph_from_data_frame(d=sw_edges, vertices=sw_nodes, directed=FALSE) # an undirected graph
g
```

Next, we output a portion of the adjacency matrix for our network.

```{r}
g[1:6, 1:6] # the first six rows and columns
```

### Visualization

Alternatively, we can show a heat map of our adjacency matrix. The number in each square equals to the weight of one edge. We observe LUKE is a very popular character.

```{r, fig.height=13, fig.width=15}
sw_matrix <- as.matrix(g[])
sw_matrix <- sw_matrix[order(rownames(sw_matrix)), order(colnames(sw_matrix))]
melted_sw_matrix <- melt(sw_matrix)
ggplot(melted_sw_matrix, aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile() +
  geom_text(aes(label=value), color="red") +
  scale_fill_gradient(low="white", high="black") +
  xlab("characters") + ylab("characters") +
  theme(axis.text.x=element_text(angle=45)) +
  labs(fill="weight")
```

We also compute characters' importance using `strength()` function based on the number of scenes they appear in and rank the importance in a descending order. The goal of `strength()` function is to sum up the edge weights of the adjacent edges for each node.

```{r}
importance <- strength(g)
sw_nodes$importance <- importance
head(arrange(sw_nodes, -importance))
```

Again, we use `visNetwork` to visualize.

```{r, fig.height=8, fig.width=15}
sw_colors <- colorRampPalette(brewer.pal(3, "RdBu"))(3)
sw_nodes$group.type <- NA
sw_nodes$group.type[sw_nodes$group=="dark side"] <- sw_colors[1]
sw_nodes$group.type[sw_nodes$group=="other"] <- sw_colors[2]
sw_nodes$group.type[sw_nodes$group=="light side"] <- sw_colors[3]
sw_nodes <- sw_nodes %>% select(-id) %>%
  mutate(id=name,
         shape="dot",
         shadow=TRUE,
         title=group,
         label=name,
         size=log((importance+3)^5), # adjust size with respect to a node's importance
         borderWidth=1,
         color.background=group.type,
         color.border="grey",
         color.highlight.background="yellow",
         color.highlight.border="black") %>% 
  arrange(id)
sw_edges <- sw_edges %>% mutate(from=source,
                                to=target,
                                width=log((weight+3)^1.5), # adjust width with respect to an edge's weight
                                color="lightgrey",
                                smooth=FALSE)
visNetwork(sw_nodes, sw_edges, width="100%", main="Star Wars Episode IV Network") %>%
  visLayout(randomSeed=21) %>% 
  visGroups(groupname="dark side", color=sw_colors[1]) %>%
  visGroups(groupname="other", color=sw_colors[2]) %>%
  visGroups(groupname="light side", color=sw_colors[3]) %>%
  visLegend(width=0.1, position="right", main="Group") %>%
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE,
             selectedBy="group") %>% 
  visInteraction(hideEdgesOnDrag=TRUE,
                 dragNodes=TRUE,
                 dragView=TRUE,
                 zoomView=TRUE,
                 navigationButtons=TRUE)
```

You may wonder how important a character is in our Star Wars network. Therefore, we want to utilize three proposed measures (**degree centrality**, **betweenness centrality**, and **closeness centrality**) to quantify each node's importance in a network and visualize how its importance is different from others.

### Centrality Measurement

Degree centrality is deﬁned as the number of adjacent edges to each node. After ranking the degree centrality, we find LUKE has the greatest value. It implies that LUKE is interacting with a great amount of unique characters. We color each node based on its degree centrality value. The node with the greatest value has the warmest color.

```{r}
degree_centrality <- degree(g)
sw_nodes$degree_centrality <- degree_centrality[as.character(sw_nodes$name)]
head(sort(degree_centrality, decreasing=TRUE))
```

```{r, fig.height=2, fig.width=15}
sw_colors_centrality <- rev(colorRampPalette(brewer.pal(9, "Oranges"))(22))
sw_nodes <- sw_nodes %>% mutate(degree_rank=23-floor(rank(degree_centrality)),
                                color.background=sw_colors_centrality[degree_rank])
network_degree <- visNetwork(sw_nodes, sw_edges, height='350px', width="100%", main="Degree Centrality") %>%
  visLayout(randomSeed=21) %>% 
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE,
             selectedBy="degree_rank") %>% 
  visInteraction(hideEdgesOnDrag=TRUE,
                 dragNodes=TRUE,
                 dragView=TRUE,
                 zoomView=TRUE,
                 navigationButtons=TRUE)
```

Betweenness centrality is deﬁned as the number of shortest paths between nodes that pass through a particular node. After ranking the betweenness centrality, we find LEIA has the greatest value. It implies that LEIA tends to be very critical to the communication process. We color each node based on its betweenness centrality value. The node with the greatest value has the warmest color.

```{r}
betweenness_centrality <- betweenness(g)
sw_nodes$betweenness_centrality <- betweenness_centrality[as.character(sw_nodes$name)]
head(sort(betweenness_centrality, decreasing=TRUE))
```

```{r, include=FALSE}
sw_nodes <- sw_nodes %>% mutate(betweenness_rank=23-floor(rank(betweenness_centrality)),
                                color.background=sw_colors_centrality[betweenness_rank])
network_betweenness <- visNetwork(sw_nodes, sw_edges, height='350px', width="100%", main="Betweenness Centrality") %>%
  visLayout(randomSeed=21) %>% 
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE,
             selectedBy="betweenness_rank") %>% 
  visInteraction(hideEdgesOnDrag=TRUE,
                 dragNodes=TRUE,
                 dragView=TRUE,
                 zoomView=TRUE,
                 navigationButtons=TRUE)
```

Closeness centrality is deﬁned as the number of steps required to access every other node from a given node. After ranking the closeness centrality, we find BIGGS has the greatest value. It implies that BIGGS is close to many other characters. We color each node based on its closeness centrality value. The node with the greatest value has the warmest color.

```{r}
closeness_centrality <- closeness(g, normalized=TRUE)
sw_nodes$closeness_centrality <- closeness_centrality[as.character(sw_nodes$name)]
head(sort(closeness_centrality, decreasing=TRUE))
```

```{r, include=FALSE}
sw_nodes <- sw_nodes %>% mutate(closeness_rank=23-floor(rank(closeness_centrality)),
                                color.background=sw_colors_centrality[closeness_rank])
network_closeness <- visNetwork(sw_nodes, sw_edges, height='350px', width="100%", main="Closeness Centrality") %>%
  visLayout(randomSeed=21) %>% 
  visOptions(highlightNearest=TRUE,
             nodesIdSelection=TRUE,
             selectedBy="closeness_rank") %>% 
  visInteraction(hideEdgesOnDrag=TRUE,
                 dragNodes=TRUE,
                 dragView=TRUE,
                 zoomView=TRUE,
                 navigationButtons=TRUE)
```

Lastly, we output our network and find discrepancies among three measurements.

```{r, echo=FALSE}
network_degree
network_betweenness
network_closeness
```

## External Resource
1. [visNetwork](https://datastorm-open.github.io/visNetwork/more.html) package;   
2. [star-wars-network](https://github.com/pablobarbera/data-science-workshop.git).