# Tidyverse Cheatsheet

Huaizhi Ge

```{r, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

## Introduction

Usually, before we plot the data, we need to preprocess the data first. And we use tidyverse to process data commonly. And I find myself can not remember all the tidyverse functions to deal with the data. So I create a cheat sheet that includes some commonly used functions in tidyverse to process the data. 

reference:
https://www.rdocumentation.org/,
https://dplyr.tidyverse.org/index.html


## select()

To select some columns in the data, we only need to add the names of columns to the Select() function.
(Use dataset "mpg" as an example dataset.)
```{r}
mpg %>% 
  select(year,manufacturer,model)
```

Also, we can use start_col:end_col.
```{r}
mpg %>% 
  select(year:drv)
```

We can select columns that have the same structure using starts_with(),end_with() or contains().

```{r}
mpg %>% 
  select(starts_with("m"),contains("c"),ends_with("y"))
```

select_if() can be used to select columns that have a specific data type. For example, select_if(is.character) can select all the character type columns. Similarly, we can use is.numeric, is.integer, is.factor and so on.

```{r}
mpg %>% 
  select_if(is.numeric)
```

select_if() can select the columns that meet some conditions.

```{r}
mpg %>% select(where(is.numeric)) %>% 
  select(where(~mean(.)<100))
```

## filter()

We can use filter() to choose certain rows.

```{r}
mpg %>% 
  filter(year==1999)
```

```{r}
mpg %>% 
  filter(year!=1999)
```

filter(condition1,condition2) will return the rows that meet the two conditions.

```{r}
mpg %>% 
  filter(year==1999,cyl==4)
```

filter(condition1|condition2) will return the rows that meet condition1 or condition2.

```{r}
mpg %>% 
  filter(year==1999|cyl==4)
```

filter(xor(condition1,condition2)) will return rows that only meet one condition.

```{r}
mpg %>% 
  filter(xor(year==1999,cyl==4))
```

filter_all() will filter all the columns.

```{r}
mpg %>% 
  filter_all(any_vars(.==4))
```

filter_if() first find the columns that meet some condition then filter.

```{r}
mpg %>% 
  filter_if(is.character,any_vars(.==4))
```

filter_at() need to select columns in vars().

```{r}
mpg %>% 
  filter_at(vars(displ),all_vars(.>3))
```

## mutate()

mutate() can add a new column based on the existing column.

```{r}
mpg %>% 
  select(manufacturer,model,year) %>% 
  mutate(year2=year-2000)
```

We can use ifelse() to choose data.

```{r}
mpg %>% 
  select(manufacturer,model,year) %>% 
  mutate(year2=ifelse(year>2000,0,NA))
```

mutate_all() can change all the columns.

```{r}
mpg %>% 
  mutate_all(toupper)
```

mutate_if() can select certain columns and mutate.

```{r}
mpg %>% 
  mutate_if(is.numeric,round)
```

mutate_at() need to select columns in vars().

```{r}
mpg %>% 
  select(manufacturer,model,displ,cyl) %>% 
  mutate_at(vars(displ,cyl),~(.*2))
```

case_when() can create categorical data.

```{r}
mpg %>% 
  select(manufacturer,model,year) %>% 
  mutate(year2=case_when(
    year<2000~"very old",
    year<2005~"old",
    TRUE~"new"
  ))
```

## rowwise()

rowwise() can process data by row.

```{r}
mpg %>% 
  select_if(is.numeric) %>% 
  rowwise() %>% 
  mutate(mean = mean(c_across(displ:hwy)))
```

Or, we can do this by another way.

```{r}
mpg %>% 
  rowwise() %>% 
  mutate(mean=rowMeans(across(where(is.numeric))))
```

We can use rowwise() to get the max/min/sum/sd too.

```{r}
mpg %>% 
  rowwise() %>% 
  mutate(sum=sum(across(where(is.numeric))))
```

## arrange()

arrange() can sort the data.

```{r}
mpg %>% 
  arrange(displ)
```

desc() can help sort data by descending order.

```{r}
mpg %>% 
  arrange(desc(displ))
```

## distinct()

distinct() can help delete the duplicate rows.

```{r}
df <- tibble(
  x=sample(5,20,rep=TRUE),
  y=sample(5,20,rep=TRUE)
)
df %>% 
  distinct()
```

## rename()

rename() can change the column name.

```{r}
mpg %>% 
  rename(YEAR=year)
```

## relocate()

relocate() can help change the order of the columns.

```{r}
mpg %>% 
  relocate(year)
```

```{r}
mpg %>% 
  relocate(year,.after = manufacturer)
```

```{r}
mpg %>% 
  relocate(year,.before = displ)
```

## drop_na()

drop_na can help drop the missing value.

```{r}
df <- tibble(
  x=c(1,NA,2,3),
  y=c(4,5,NA,6)
)
df %>% 
  drop_na()
```

## pull()

pull() can help get a single column.

```{r}
mpg %>% 
  pull(year)
```

## unite()

unite() can help unite the columns.

```{r}
mpg %>% 
  unite(col="newcolumn",manufacturer:model,sep=",")->newdf

newdf
```

## separate()

separate() can help separate one column to several columns.

```{r}
newdf %>% 
  separate("newcolumn",into=c("manufacturer","model"),sep=",")
```

## pivot_longer()

pivot_longer() "lengthens" data, increasing the number of rows and decreasing the number of columns.

```{r}
df <- tibble(index=c(1,2),x=c('a','b'),y=c('c','d'),z=c('e','f'))
df
```

```{r}
df2 <- df %>% 
  pivot_longer(cols=!index,names_to="key",values_to="val")
df2
```

## pivot_wider()

pivot_wider() "widens" data, increasing the number of columns and decreasing the number of rows.

```{r}
df2 %>% 
  pivot_wider(names_from=key,values_from=val)
```

## count()

count() can count the number of occurences.

```{r}
mpg %>% 
  count(model,sort=TRUE)
```

## summarize()

summarise() is typically used on grouped data created by group_by(). The output will have one row for each group.

```{r}
mpg %>% 
  group_by(model) %>% 
  summarise(count=n(),averagedispl=mean(displ),maxdispl=max(displ))
```

summarise_all() applies the functions to all (non-grouping) columns.

```{r}
mpg %>% 
  group_by(model) %>% 
  select_if(is.numeric) %>% 
  summarise_all(mean)
```

summarise_if() operate on columns for which a predicate returns TRUE.

```{r}
mpg %>% 
  group_by(model) %>% 
  summarise_if(is.numeric,mean)
```

summarise_at() allow you to select columns.

```{r}
mpg %>% 
  group_by(model) %>% 
  summarise_at(vars(displ,cyl),mean)
```

