
#Using PostgreSQL Databse in R with MacOS Environment
wl2824 Wei Luo

### 1. Requirements

The required tools and packages are:

            1. R installation (assumed)
            2. PostgreSQL Database installation
            3. RMariaDB package
            
### 1.2. PostgreSQL installation guide with HomeBrew


Open a terminal in your laptop and run the following command:

    1. /bin/bash -c "$(curl -fsSL https://raw.githubusercontent.com/Homebrew/install/HEAD/install.sh)"

This will install homebrew service to your mac.(You will see a Updating Homebrew message after running this command. If it went too long, you can press Crtl + c to stop.)

After successfully installed homebrew:

    2. sudo brew install postgresql

To confirm that Postgres was installed correctly, run the following command and make sure that it outputs version information (not an error message):

    3. psql --version
    
    
### 1.3. PostgreSQL setup 
    
To start the psql server:

    1. pg_ctl -D /usr/local/var/postgres start
    
To start postgres, type the command in terminal: 

    2. psql postgres
    
You can create a new Database with PostgreSQl using command:

    3. CREATE DATABASE name_of_your_choice;

### 2.1 Connect to PostGreSQL Database in R Studio

Required R packages:

    1. odbc
    2. DBI
    3. dplyr
    4. pool

In order to Connect to a PostgreSQL database, run the following command in R studio:

```{r}

library(DBI)
library(dplyr)
library(ggplot2)

# SQLite database
con <- dbConnect(RSQLite::SQLite(), ":memory:")
#copy_to(con, nycflights13::flights, "FLIGHTS")

# ODBC databases (requires a live database connection)
#con <- dbConnect(odbc::odbc(), "SQL Server")
#con <- dbConnect(odbc::odbc(), "Snowflake")


```

You can also run SQL queries with R studio after connecting to PostgreSQL Database:
```{r}
    #data <- dbSendQuery(con, "your query")
    #dbBind(data, list("col A", "MSY"))
    #dbFetch(airport)
```
Then all the data fetched from the PostgreSQL will be in data. You can do any data visualization with the dataset using R. 

### 3. Walking through an example.

In this section, we are going to use a simple csv file as our initial dataset, upload it to the PostgreSQL database and then fetch it using R in R studio.

In the attachment, you will find an example csv file called "part.csv"

Create a file called "loaddata.sql":

    BEGIN;
    COPY part FROM 'change it to your absolute path/part.csv' WITH (FORMAT csv, DELIMITER '|');
    COMMIT;

Then create a file called "ddl.sql":

    DROP TABLE IF EXISTS part CASCADE;

    CREATE TABLE  part ( p_partkey integer, p_1 integer,  p_2 integer, p_3 integer, p_4 integer, p_5 integer, p_6 integer, p_7 integer, p_8 integer, primary key (p_partkey));
    
Create a file called "setup.sh":

    CUR_PATH=`pwd`

    psql yourdatabase < ${CUR_PATH}/ddl.sql
    echo "Created tables"

    psql yourdatabase < ${CUR_PATH}/loaddata.sql
    echo "Loaded data into tables"

Before you run setup.sh, please change corresponding path in loaddata.sql.

Run setup.sh by ./setup.sh or bash setup.sh, then your PostgreSQL should fill with the form called "part".

Configure your DB connection details:

```{r}
#con <- DBI::dbConnect(odbc::odbc(),
#                      Driver   = "[your driver's name]",
#                      Server   = "[your server's path]",
#                      Database = "[your database's name]",
#                      UID      = rstudioapi::askForPassword("Database user"),
#                      PWD      = rstudioapi::askForPassword("Database #password"),
#                      Port     = 5432)
```

Then you can connect to database and use SQL queries to find whatever you want!
